---
title: "Fitting delta models with sdmTMB including separate models and built-in function"
author: "Julia Indivero, Sean Anderson, Lewis Barnett, Philina English, Eric Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting delta models with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
inla_installed <- requireNamespace("INLA", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && inla_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

sdmTMB has the capability for built-in hurdle models (also called delta models). 
Hurdle models could also be implemented by fitting the two components separately and combining the predictions.

Hurdle models are more appropriate than something like a Tweedie when there are differences in the processes controlling presence vs. abundance, or when greater flexibility to account for dispersion is required. 

Built-in hurdle models can be specified with the `family` argument within the `sdmTMB()` function. Current options include:

1. Delta-Gamma: `family = delta_gamma(link1 = "logit", link2 = "log")`.
This fits a binomial presence-absence model (i.e., `family = binomial(link = "logit")`) and then a model for the positive catches only with a Gamma observation distribution and a log link (i.e., `family = Gamma(link = "log")`). Here and with other delta models, the `link1` and `link2` can be omitted and left at their default values.

2. Delta-lognormal: `family = delta_lognormal()`.
This fits a binomial presence-absence model (i.e., `family = binomial(link = "logit")`) and then a model for the positive catches only with a lognormal observation distribution and a log link (i.e., `family = lognormal(link = "log")`

<!-- Currently not supported: 3. A Poisson-link delta model: `family = delta_poisson_link_gamma()` -->
<!-- This fits a binomial presence-absence model (`family = binomial(link = "logit")`) and then a model for the positive catches only with a Poisson observation distribution and a log link  -->

3. Delta-truncated-negative-binomial: `family = delta_truncated_nbinom2()` or `family = delta_truncated_nbinom2()`.
This fits a binomial presence-absence model (`family = binomial(link = "logit")`) and a `family = truncated_nbinom2(link = "log")` or `family = truncated_nbinom1(link = "log")` distribution for positive catches.

To summarize the built-in delta models and the separate components:
Model Type         |Built-in delta function              | Presence-absence model            | Positive catch model                    |
-------------------|-------------------------------------|-----------------------------------|-----------------------------------------|
Delta-gamma        |`family = delta_gamma()`             |`family = binomial(link = "logit")`| `family = Gamma(link = "log")`          |
Delta-lognormal    |`family = delta_lognormal()`         |`family = binomial(link = "logit")`| `family = lognormal(link = "log")`      |
Delta-NB           |`family = delta_truncated_nbinom2()` |`family = binomial(link = "logit")`|`family = truncated_nbinom(link = "log")`|


<!-- Add back to table when included again: Poisson-link delta |`family = delta_poisson_link_gamma()`|`family = binomial(link = "logit")`| `family = lognormal(link = "log")`  -->

# Example of fitting with built-in delta model

Here, we will show an example of fitting each of the delta model types with the built-in features, as well as how to build each model component separately and then combine.

We will use a dataset built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound, British Columbia, Canada. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

We will first create a mesh that we will use for all the models.

```{r}
pcod_mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

Then we can fit a model of Pacific cod density using a delta-gamma model, including a smoothed effect of depth.

```{r, warning=F, message=F}
 fit_dg <- sdmTMB(density ~ 1 + s(depth),
                     data = pcod, 
                    mesh = pcod_mesh,
                    time = "year", 
                    family = delta_gamma()
    )
```

The default in built-in delta models is for the formula, spatial and spatio-temporal structure, and anisotropy to be shared between the two model components. However, some elements (`formula`, `spatial`, `spatiotemporal`, and `share_range`) can also be specified independently for each model using a list format within the function argument (see examples below). The first element of the list is for the binomial component and the second element is for the positive component (e.g., Gamma). 

Some elements must be shared for now (e.g., spatially varying coefficients, and time-varying coefficients). 

To specify the settings for spatial and spatio-temporal effects in each model component, create a list of settings within the `spatial` and `spatiotemporal` arguments, i.e., `spatial = list("setting for presence-absence model, "setting for positive catch model")`.
For example, we could turn spatial and spatiotemporal effects off for the presence-absence model but include spatial effects for the positive catch model and include spatiotemporal effects as a random walk.

```{r, warning=F, message=F, echo=F}
fit_dg2 <- sdmTMB(density ~ 1 + s(depth),
  data = pcod, 
  mesh = pcod_mesh,
  time = "year",
  spatial = list("off", "on"), #<
  spatiotemporal = list("off", "rw"), #<
  family = delta_gamma()
)
```

We could similarly specify a different formula for each component of the model, using `list(formula for model 1, formula for model 2)`
For instance, we could include the effect of depth for only the positive catch model, and remove it for the presence-absence model.

However, there are currently limitations if specifying separate formulas for each model component. The two formulas *cannot* have: 
* smoothers
* threshold effects
* random intercepts
For now, these must be specified through a single formula that is shared across the two models. 

```{r, warning=F, message=F, echo=F, results='hide'}
fit_dg3 <- sdmTMB(
  list(
    density ~ 1, #<
    density ~ poly(depth, 2) #<
  ),
  data = pcod, 
  mesh = pcod_mesh,
  time = "year",
  spatial = list("off", "on"),
  spatiotemporal = list("off", "rw"),
  family = delta_gamma()
)
```

Each model component can similarly have separate settings for `share_range`, which determines whether there is a shared spatial and spatiotemporal range parameter (`TRUE`) or independent range parameters (`FALSE`), by using a list.

```{r, warning=F, message=F, echo=F, results='hide'}
 fit_dg4 <- sdmTMB(density ~ 1 + s(depth),
                     data = pcod, 
                    mesh = pcod_mesh,
                    time = "year", 
                    share_range=list(TRUE, FALSE), #<
                    family = delta_gamma()
    )
```

Lastly, whether or not anisotropy is included in the model is determined with the logical argument `anisotropy=()` (i.e. `TRUE` or `FALSE`), and cannot be separately specified for each model. If anisotropy is included, it is by default shared across the two model components. However it can be made unique in each model component by using `sdmTMBcontrol()` and adding the argument `control=()` when fitting the model.

```{r, warning=F, message=F, echo=F, results='hide}
 fit_dg5 <- sdmTMB(density ~ 1 + s(depth),
                     data = pcod, 
                    mesh = pcod_mesh,
                    time = "year", 
                   control = sdmTMBcontrol(map = list(ln_H_input = factor(c(1, 2, 3, 4)))), #<
                    family = delta_gamma()
    )
```

Once we fit the delta model, we can evaluate and plot the output, diagnostics, and predictions similar to other sdmTMB models.

The printed model output will show estimates and standard errors of parameters for each model separately.
```{r}
print(fit_dg)
```

Using the `tidy()` function will turn the sdmTMB model output into a data frame, with the argument `model=1` or `model=2` to specify which model component to extract as a dataframe. See [`?tidy.sdmTMB`](https://pbs-assess.github.io/sdmTMB/reference/tidy.sdmTMB.html) for additional arguments and options.
```{r}
fit_dg_output <- tidy(fit_dg, model=1)
```

We can use `predict()` to predict density in either log space (the default) or transformed into the response scale (`type = "response"`). 

For built-in delta models, the default function will return estimated response and parameters for each grid cell for each model separately, notated with a 1 (for the presence/absence model) or 2 (for the positive catch model) in the column name. See [`?predict.sdmTMB`](https://pbs-assess.github.io/sdmTMB/reference/tidy.sdmTMB.html) for description of values in the dataframe.
```{r, error=F, message=F}
p <- predict(fit_dg, newdata = qcs_grid)
p_response <- predict(fit_dg, newdata = qcs_grid, type = "response")
head(p)
```

We can use the `model=()` argument to return only predictions from model 1 or model 2.
```{r, error=F, message=F, echo=F}
p1 <- predict(fit_dg, newdata = qcs_grid, model=1)
p2 <- predict(fit_dg, newdata = qcs_grid, model=2)
```

To return combined predictions from both components on the response scale, we can use `model=NA`
```{r, error=F, message=F, echo=F}
p_combined <- predict(fit_dg, newdata = qcs_grid, model=NA, type="response")
```

We can use predictions from the built-in delta model (making sure that `return_tmb_object=T`) to get the index values using the `get_index()` function. This can be used with predictions that include both the first and second models (i.e. using the default and specifying no model=() argument) or with predictions generated using `model=NA`. The get_index() function will automatically combine predictions from the first and second model in calculating the index values. For more on modelling for the purposes of creating an index see the vignette on [Index standardization with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/index-standardization.html).

```{r, message=F, warning=F}
p2 <- predict(fit_dg, newdata = qcs_grid, return_tmb_object=T)
ind_dg <- get_index(p2, bias_correct = FALSE)
head(ind_dg)
```

We can plot conditional effects of covariates (such as depth in the example model) using the package `visreg` by specifying each model component with `model=1` for the presence-absence model or `model=2` for the positive catch model. Currently, plotting effects of built-in delta models with `ggeffects` is not supported. See the vignette on using [visreg](link) with sdmTMB for more information. 
```{r}
visreg_delta(fit_dg, xvar = "depth", model = 1, gg = TRUE)
visreg_delta(fit_dg, xvar = "depth", model = 2, gg = TRUE)
```
<!-- TODO Will it work with ggeffects package? -->

The built-in delta models can also be evaluated with the `residuals()` functions in sdmTMB. Similarly to generating predictions, we can specify which of the model components we want to return residuals for using the `model=()` argument and specifying `=1` or `=2`. See [documentation here] (https://rdrr.io/github/pbs-assess/sdmTMB/man/residuals.sdmTMB.html) for additional options for evaluating residuals in sdmTMB models. 
```{r, echo=F, message=F, results='hide, warning=F}
  residuals_1 <- residuals(fit_dg,type="response", model=1)
  residuals_2 <- residuals(fit_dg,type="response", model=2)
```

We can also simulate new observations from a fitted delta model. As in other functions, we can specify which model to simulate from using the argument `model=1` for only presence/absence, `model=2` for only positive catches, or `model=NA` for combined predictions. See [`?simulate.sdmTMB`](https://rdrr.io/github/pbs-assess/sdmTMB/man/simulate.sdmTMB.html) for more details on simulation options.

```{r, message=F, warning=F}
simulations <- simulate(fit_dg, nsim=5, seed=5090, model=NA)
```

# Delta models by fitting two components separately and combining predictions
Next, we will show an example of how to implement a delta-gamma model in sdmTMB, but with each component fit separately and then combined.

```{r glimpse-pcod}
glimpse(pcod)
```

```{r spde, fig.asp=0.7}
mesh1 <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

It is not necessary to use the same mesh for both models, but one can do so by updating the first mesh to match the reduced data frame as shown here:

```{r, fig.asp=0.9}
dat2 <- subset(pcod, density > 0)
mesh2 <- make_mesh(dat2,
  xy_cols = c("X", "Y"),
  mesh = mesh1$mesh
)
plot(mesh2)
```

This delta-gamma model is similar to the Tweedie model in the [Intro to modelling with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/basic-intro.html) vignette, except that we will use `s()` for the depth effect. 

```{r model1, warning=FALSE}
m1 <- sdmTMB(
  formula = present ~ 0 + as.factor(year) + s(depth, k = 3),
  data = pcod,
  mesh = mesh1,
  time = "year", family = binomial(link = "logit"),
  spatiotemporal = "iid",
  spatial = "on"
)
m1
```

One can use different covariates in each model, but in this case we will just let the depth effect be more wiggly by not specifying `k = 3`. 
```{r model2, warning=FALSE}
m2 <- sdmTMB(
  formula = density ~ 0 + as.factor(year) + s(depth),
  data = dat2,
  mesh = mesh2,
  time = "year",
  family = Gamma(link = "log"),
  spatiotemporal = "iid",
  spatial = "on"
)
m2
```

We can inspect the effect of the smoothed covariate using `plot_smooth()`.
```{r model1-depth}
s1 <- plot_smooth(m1, ggplot = TRUE)
s1 + xlim(min(pcod$depth), max(pcod$depth))
```

```{r model2-depth}
s2 <- plot_smooth(m2, ggplot = TRUE)
s2 + xlim(min(pcod$depth), max(pcod$depth))
```

Next, we need some way of combining the predictions across the two models. 
If all we need are point predictions, we can just multiply the predictions from the two models after applying the inverse link:

```{r delta-pred-simple, echo=TRUE}
pred <- qcs_grid # use the grid as template for saving our predictions
p_bin <- predict(m1, newdata = qcs_grid)
p_pos <- predict(m2, newdata = qcs_grid)
p_bin_prob <- m1$family$linkinv(p_bin$est)
p_pos_exp <- m2$family$linkinv(p_pos$est)
pred$est_exp <- p_bin_prob * p_pos_exp
```

But if a measure of uncertainty is required, we can simulate from the joint parameter precision matrix using the `predict()` function with any number of simulations selected (e.g., `sims = 500`). 
Because the predictions come from simulated draws from the parameter covariance matrix, the predictions will become more consistent with a larger number of draws. 
However, a greater number of draws takes longer to calculate and will use more memory (larger matrix), so fewer draws (~100) may be fine for experimentation. 
A larger number (say ~1000) may be appropriate for final model runs.

```{r delta-pred-sim, echo=TRUE}
set.seed(28239)
p_bin_sim <- predict(m1, newdata = qcs_grid, nsim = 500)
p_pos_sim <- predict(m2, newdata = qcs_grid, nsim = 500)
p_bin_prob_sim <- m1$family$linkinv(p_bin_sim)
p_pos_exp_sim <- m2$family$linkinv(p_pos_sim)
p_combined_sim <- p_bin_prob_sim * p_pos_exp_sim
```

`p_combined_sim` is just a matrix with a row for each row of data that was predicted on and width `nsim`. 
You can process this matrix however you would like.
We can save median predictions and upper and lower 95% confidence intervals:

```{r, fig.width=5}
pred$median <- apply(p_combined_sim, 1, median)
# pred$lwr <- apply(p_combined_sim, 1, quantile, probs = 0.025)
# pred$upr <- apply(p_combined_sim, 1, quantile, probs = 0.975)
plot(pred$est_exp, pred$median)
```

```{r}
ggplot(subset(pred, year == 2017), aes(X, Y, fill = median)) +
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "sqrt")
```

And we can now calculate spatial uncertainty:

```{r cv}
pred$cv <- apply(p_combined_sim, 1, function(x) sd(x) / mean(x))
ggplot(subset(pred, year == 2017), aes(X, Y, fill = cv)) + # 2017 as an example
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10")
```

sdmTMB also has a function for calculating an index from those draws `get_index_sims()`.
This function is just summing biomass or abundance across grid cells for each simulation draw and for each year and then calculating quantiles on the distribution of samples.
The default for this function expects the simulations to still be in log space, so we either need to `log(p_combined)`, or we can set `agg_function = function(x) sum(x)`. 

```{r get_index}
qcs_grid$area <- 4 # all 2 x 2km
ind <- get_index_sims(p_combined_sim / 1000, # convert from kg to tonnes
  agg_function = function(x) sum(x),
  area = qcs_grid$area
)
ggplot(ind, aes(year, est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  ylab("Biomass (t)")
```
