---
title: "Fitting delta models with sdmTMB including separate models and built-in function"
author: "Julia Indivero, Sean Anderson, Lewis Barnett, Philina English, Eric Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting delta models with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
inla_installed <- requireNamespace("INLA", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && inla_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

sdmTMB has the capability for built-in hurdle models (also called delta models). Hurdle models could also be implemented by fitting the two components separately and combining the predictions.

Hurdle models are more appropriate than something like a Tweedie when there are differences in the processes controlling presence vs. abundance, or when greater flexibility to account for dispersion is required. 

Built-in hurdle models can be specified in the `family=()` argument within the `sdmTMB()` function. The options include:

1. Delta-Gamma model: `family=delta_gamma()`  
This fits a binomial presence-absence model (ie `family = binomial(link = "logit")`) and then a model for the positive catches only with a Gamma observation distribution and a log link (ie `family= Gamma(link = "log")`)

2.A delta-lognormal model: `family=delta_lognormal()`
This fits a binomial presence-absence model (ie `family = binomial(link = "logit")`) and then a model for the positive catches only with a lognormal observation distribution and a log link (ie `family= lognormal(link = "log")`

3. A Poisson-link delta model: `family=delta_poisson_link_gamma()`
This fits a binomial presence-absence model (`family = binomial(link = "logit")`) and then a model for the positive catches only with a Poisson observation distribution and a log link 

4. Delta-binomial model: `family=delta_truncated_nbinom2()`
This fits a binomial presence-absence model (`family = binomial(link = "logit")`) and zero-inflated count data and a `family=truncated_nbinom(link = "log")` or `family=truncated_nbinom2(link = "log")` distribution for positive catches

To summarize the built-in delta models and the separate components:
Model Type         |Built-in delta function            | Presence-absence model            | Positive catch model                  |
-------------------|-----------------------------------|-----------------------------------|---------------------------------------|
Delta-gamma        |`family=delta_gamma()`             |`family = binomial(link = "logit")`| `family = Gamma(link = "log")`        |
Delta-lognormal    |`family=delta_lognormal()`         |`family = binomial(link = "logit")`| `family = lognormal(link = "log")`    |
Poisson-link delta |`family=delta_poisson_link_gamma()`|`family = binomial(link = "logit")`| `family = lognormal(link = "log")` n  |
Delta-binomial     |`family=delta_truncated_nbinom2()` |`family = binomial(link = "logit")`|`family=truncated_nbinom(link = "log")`|


# Example of fitting with built-in delta model
Here, we will show an example of fitting each of the delta model types with the built-in features, as well as how to build each model component separately and then combine.

We will use a dataset built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound, British Columbia, Canada. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

We will first create a mesh that we will use for all the models.
```{r}
pcod_mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

Then we can fit a model of Pacific cod density using a delta-gamma model, including a smoothed effect of depth.
```{r}
 fit_dg <- sdmTMB(density ~ 1 + s(depth),
                     data = pcod, 
                    mesh = pcod_mesh,
                    time = "year", 
                    family = delta_gamma()
    )
```

To specify the settings for spatial and spatio-temporal effects in each model component, create a list of settings within the `spatial=()` and `spatiotemporal=()` arguments, ie `spatial=list("setting for presence-absence model, "setting for positive catch model")`. For example, we could turn spatial and spatiotemporal effects off for the presence-absence model but include spatial effects for the positive catch model and include spatio-temporal effects as a random walk. See 

```{r}
 fit_dg2 <- sdmTMB(density ~ 1 + s(depth),
                     data = pcod, 
                    mesh = pcod_mesh,
                    time = "year",
                  spatial = list("off", "on"),
                  spatiotemporal = list("off", "rw"),
                    family = delta_gamma()
    )

```

We could similarly specify a different equation for each component of the model. For instance, we could include the effect of depth for only the positive catch model, and remove it for the presence-absence model.

```{r}

```

We can then predict density in either log space or transformed into the response scale.
```{r}
p <- predict(fit_dg, newdata = qcs_grid)
p_response <- predict(fit_dg, newdata = qcs_grid, type = "response")
```

We can use predictions to get the index. For more on modelling for the purposes of creating an index see the vignette on [Index standardization with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/index-standardization.html).
```{r}
ind_dg <- get_index(p, bias_correct = FALSE)
```

We can plot the effect of depth using the package `visreg` by specifying each model component with `model=1` for the presence-absence model or `model=2`for the positive catch model. See the vignettes on using [visreg](link) and [ggeffects](link) for more information about plotting effects from sdmTMB models.
```{r}
visreg_delta(fit_dg, xvar = "depth", model = 1, gg = TRUE)
visreg_delta(fit_dg, xvar = "depth", model = 2, gg = TRUE)

#Will it work with ggeffects package?
```

The built-in delta models can also be evaluated with the `residuals` functions in sdmTMB. See [documentation here] (https://rdrr.io/github/pbs-assess/sdmTMB/man/residuals.sdmTMB.html) for additional options for residuals in sdmTMB
```{r, echo=F, message=F, results='hide, warning=F}
  set.seed(2938)
  residuals <- residuals(fit_dg, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
```

```{r}
  qqnorm(residuals)
  qqline(residuals)
```

We can also simulate data from a built-in delta model
```{r}
#Need help with how to simulate data from the delta model
set.seed(1)
predictor_dat <- data.frame(X = runif(300), Y = runif(300))
mesh <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), cutoff = 0.1)
dat <- sdmTMB_simulate(
  formula = ~ 1,
  data = predictor_dat,
  mesh = mesh,
  family = delta_gamma(),
  range = 0.5,
  sigma_O = 0.2,
  seed = 42,
  B = c(0.2)
)

fit <- sdmTMB(observed ~ 1, data = dat, family = delta_gamma(), mesh = mesh)

# simulate from the model:
s1 <- simulate(fit, nsim = 300)
dim(s1)
```

# Delta models by fitting two components separately and combining predictions
Next, we will show an example of how to implement a delta-gamma model in sdmTMB, but with each component fit separately and then combined.

```{r glimpse-pcod}
glimpse(pcod)
```

```{r spde, fig.asp=0.7}
mesh1 <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

It is not necessary to use the same mesh for both models, but one can do so by updating the first mesh to match the reduced data frame as shown here:

```{r, fig.asp=0.9}
dat2 <- subset(pcod, density > 0)
mesh2 <- make_mesh(dat2,
  xy_cols = c("X", "Y"),
  mesh = mesh1$mesh
)
plot(mesh2)
```

This delta-gamma model is similar to the Tweedie model in the [Intro to modelling with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/basic-intro.html) vignette, except that we will use `s()` for the depth effect. 

```{r model1, warning=FALSE}
m1 <- sdmTMB(
  formula = present ~ 0 + as.factor(year) + s(depth, k = 3),
  data = pcod,
  mesh = mesh1,
  time = "year", family = binomial(link = "logit"),
  spatiotemporal = "iid",
  spatial = "on"
)
m1
```

One can use different covariates in each model, but in this case we will just let the depth effect be more wiggly by not specifying `k = 3`. 
```{r model2, warning=FALSE}
m2 <- sdmTMB(
  formula = density ~ 0 + as.factor(year) + s(depth),
  data = dat2,
  mesh = mesh2,
  time = "year",
  family = Gamma(link = "log"),
  spatiotemporal = "iid",
  spatial = "on"
)
m2
```

We can inspect the effect of the smoothed covariate using `plot_smooth()`.
```{r model1-depth}
s1 <- plot_smooth(m1, ggplot = TRUE)
s1 + xlim(min(pcod$depth), max(pcod$depth))
```

```{r model2-depth}
s2 <- plot_smooth(m2, ggplot = TRUE)
s2 + xlim(min(pcod$depth), max(pcod$depth))
```

Next, we need some way of combining the predictions across the two models. 
If all we need are point predictions, we can just multiply the predictions from the two models after applying the inverse link:

```{r delta-pred-simple, echo=TRUE}
pred <- qcs_grid # use the grid as template for saving our predictions
p_bin <- predict(m1, newdata = qcs_grid)
p_pos <- predict(m2, newdata = qcs_grid)
p_bin_prob <- m1$family$linkinv(p_bin$est)
p_pos_exp <- m2$family$linkinv(p_pos$est)
pred$est_exp <- p_bin_prob * p_pos_exp
```

But if a measure of uncertainty is required, we can simulate from the joint parameter precision matrix using the `predict()` function with any number of simulations selected (e.g., `sims = 500`). 
Because the predictions come from simulated draws from the parameter covariance matrix, the predictions will become more consistent with a larger number of draws. 
However, a greater number of draws takes longer to calculate and will use more memory (larger matrix), so fewer draws (~100) may be fine for experimentation. 
A larger number (say ~1000) may be appropriate for final model runs.

```{r delta-pred-sim, echo=TRUE}
set.seed(28239)
p_bin_sim <- predict(m1, newdata = qcs_grid, nsim = 500)
p_pos_sim <- predict(m2, newdata = qcs_grid, nsim = 500)
p_bin_prob_sim <- m1$family$linkinv(p_bin_sim)
p_pos_exp_sim <- m2$family$linkinv(p_pos_sim)
p_combined_sim <- p_bin_prob_sim * p_pos_exp_sim
```

`p_combined_sim` is just a matrix with a row for each row of data that was predicted on and width `nsim`. 
You can process this matrix however you would like.
We can save median predictions and upper and lower 95% confidence intervals:

```{r, fig.width=5}
pred$median <- apply(p_combined_sim, 1, median)
# pred$lwr <- apply(p_combined_sim, 1, quantile, probs = 0.025)
# pred$upr <- apply(p_combined_sim, 1, quantile, probs = 0.975)
plot(pred$est_exp, pred$median)
```

```{r}
ggplot(subset(pred, year == 2017), aes(X, Y, fill = median)) +
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "sqrt")
```

And we can now calculate spatial uncertainty:

```{r cv}
pred$cv <- apply(p_combined_sim, 1, function(x) sd(x) / mean(x))
ggplot(subset(pred, year == 2017), aes(X, Y, fill = cv)) + # 2017 as an example
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10")
```

sdmTMB also has a function for calculating an index from those draws `get_index_sims()`.
This function is just summing biomass or abundance across grid cells for each simulation draw and for each year and then calculating quantiles on the distribution of samples.
The default for this function expects the simulations to still be in log space, so we either need to `log(p_combined)`, or we can set `agg_function = function(x) sum(x)`. 

```{r get_index}
qcs_grid$area <- 4 # all 2 x 2km
ind <- get_index_sims(p_combined_sim / 1000, # convert from kg to tonnes
  agg_function = function(x) sum(x),
  area = qcs_grid$area
)
ggplot(ind, aes(year, est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  ylab("Biomass (t)")
```






```{r, eval=F, include=F}
#Other types of built-in delta models if we want to include
fit_dln <- sdmTMB(density ~ 1,
                      data = pcod, mesh = pcod_spde,
                      spatial = "off",
                      family = delta_lognormal())
fit_plg <- sdmTMB(density ~ 1,
                      data = pcod, mesh = pcod_spde,
                      spatial = "off",
                      family = delta_poisson_link_gamma())
  fit_dtnb2 <- sdmTMB(count ~ 1,
                        data = pcod, mesh = pcod_spde, spatial = "off",
                        family = delta_truncated_nbinom2())

```

