---
title: "Forecasting with sdmTMB"
author: "Julia Indivero, Sean Anderson, Lewis Barnett, Philina English, Eric Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
visreg_installed <- require("visreg", quietly=TRUE)
sdmTMB_installed <- require("sdmTMB", quietly=TRUE)
pkgs <- dplyr_installed && ggplot_installed && visreg_installed && sdmTMB_installed

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

```{r echo=FALSE}
library(sdmTMB)
library(ggplot2)
library(visreg)
```

---
title: "Forecasting with sdmTMB"
subtitle: "NOAA PSAW Seminar Series"
author: ""
institute: ""
date: "March 9, 2022"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("noaa-psaw-2022/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# Types of prediction we might be interested in 

* Time  
  * e.g., interpolating over missed survey year
  * e.g., forecasting future year
  
* Extrapolating in space
  * e.g., unsampled area (MPA, area beyond existing domain)

---

class: center, middle, inverse

# Forecasting in time

---


# Predicting to missing/future years

* Need a model for time: e.g., can't predict with years as factors

* Options: 
  * AR(1) or random walk random fields
  * Time-varying intercept
  * Smoother on year (`s(year)`)
  * Ignore time (fixed)
  * Some combination of these

---

# AR(1) spatiotemporal field

```{r pcod-mesh}
mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

.small[
```{r fit-ar1, echo=TRUE, eval=TRUE, results='hide'}
# missing and forecasted years:
extra_years <- c(
  2006, 2008, 2010, 2012, 2014, 2016,
  2018:2025
)
fit_ar1 <- sdmTMB(
  density ~ depth_scaled + depth_scaled2,
  time = "year",
  extra_time = extra_years, #<<
  spatiotemporal = "AR1", #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "off",
  silent = FALSE
)
```
]

---

# Random walk spatiotemporal field

.small[
```{r fit-rw, echo=TRUE, eval=TRUE, results='hide'}
fit_rw <- sdmTMB(
  density ~ depth_scaled + depth_scaled2,
  time = "year",
  extra_time = extra_years, #<<
  spatiotemporal = "RW", #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "off",
  silent = FALSE
)
```
]

---

# Random walk intercept + AR(1) fields

.small[
```{r, fit-rw-ar1, echo=TRUE, eval=TRUE, results='hide'}
fit_rw_ar1 <- sdmTMB(
  density ~ 0 + depth_scaled + depth_scaled2,
  time = "year",
  time_varying = ~1, #<<
  extra_time = extra_years, #<<
  spatiotemporal = "AR1", #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "off",
  silent = FALSE
)
```
]

---

# Smoother on year + AR(1) fields

.small[
```{r, fit-sm, echo=TRUE, eval=TRUE, results='hide'}
fit_sm <- sdmTMB(
  density ~ s(year) + depth_scaled + depth_scaled2, #<<
  time = "year",
  extra_time = extra_years, #<<
  spatiotemporal = "AR1", #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "off",
  silent = FALSE
)
```
]

---

# Comparing predicted density at a point in space

.xsmall[Vertical dashed lines indicate observations]

```{r pred-all}
newdf <- data.frame(
  year = unique(fit_ar1$data$year),
  X = mean(pcod$X),
  Y = mean(pcod$Y),
  depth_scaled = mean(pcod$depth_scaled),
  depth_scaled2 = mean(pcod$depth_scaled)^2
)
fits <- list(fit_ar1, fit_rw, fit_rw_ar1, fit_sm)
names(fits) <- c("AR1-fields", "RW-fields", "RW-time, AR1-fields", "s(year), AR1-fields")
set.seed(123)
preds <- purrr::map(fits, ~ predict(.x, newdf, nsim = 50L))
preds_df <- purrr::map_dfr(preds, function(.x) {
  reshape2::melt(.x, value.name = "est") %>%
    rename(year = Var1, iter = Var2)
}, .id = "type")
pred_df <- purrr::map_dfr(fits, ~ predict(.x, newdf, se_fit = TRUE), .id = "type")
```

```{r plot-time-comparison, fig.asp=0.62}
set.seed(1922)
iters <- sample(1:max(preds_df$iter), 12L)
ggplot(pred_df, aes(year, exp(est), ymin = exp(est - 2 * est_se), ymax = exp(est + 2 * est_se))) +
  geom_ribbon(alpha = 0.2) +
  geom_line(lwd = 1) +
  scale_y_log10() +
  facet_wrap(vars(type)) +
  geom_vline(xintercept = unique(pcod$year), lty = 2, alpha = 0.2) +
  theme(panel.grid = element_blank()) +
  geom_line(aes(year, exp(est), group = iter),
    alpha = 0.4,
    data = filter(preds_df, iter %in% iters), inherit.aes = FALSE, lwd = 0.5
  ) +
  ylab("Fish density (log-distributed axis)") + xlab("")
```

---

# AR(1) spatiotemporal fields evolve towards mean zero

```{r pred-ar1-plot-eps}
one_yr <- dplyr::filter(qcs_grid, year == 2017)
grid <- purrr::map_dfr(unique(fit_ar1$data$year), function(i) {
  one_yr$year <- i
  one_yr
})
p_ar1 <- predict(fit_ar1, newdata = grid)
ggplot(p_ar1, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed()
```

---

# Random walk fields do not evolve towards mean

```{r pred-rw-plot-eps}
p_rw <- predict(fit_rw, newdata = grid)
ggplot(p_rw, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed()
```

---

# Spatiotemporal field uncertainty grows without data

(Here AR(1) fields; random walk similar.)

```{r eps-se}
eps_est <- predict(fit_ar1, newdata = grid, nsim = 100, sims_var = "proj_re_st_vector")
grid$se <- apply(eps_est, 1, sd)
ggplot(grid, aes(X, Y, fill = se)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  labs(fill = "Standard error\nof spatiotemporal field")
```

---

# Forecasting and interpolating in time summary

* Use `extra_time` argument to fill in or forecast  
  .xsmall[(a more flexible interface for forecasting may be forthcoming)]
* Need a model for time
* AR(1) field processes revert towards mean
* Random walk field processes do not revert towards mean
* Smoothers should be used with caution when forecasting  
  .xsmall[(they continue whatever the basis functions were doing)]

---

class: center, middle, inverse

# Spatial extrapolation

---

# Previously explored BEI tree dataset

```{r bei, echo=FALSE, eval=TRUE}
dat <- data.frame(
  x = spatstat.data::bei$x,
  y = spatstat.data::bei$y
)
ggplot(dat, aes(x, y)) +
  geom_point(col = "darkblue", alpha = 0.1) +
  coord_cartesian(expand = FALSE)
```

---

# Mesh more uncertain in unsampled areas

```{r bei2}
# scale controls resolution
scale <- 50
dat$x <- scale * floor(dat$x / scale)
dat$y <- scale * floor(dat$y / scale)

dat <- dplyr::group_by(dat, x, y) %>%
  dplyr::summarise(n = n())

mesh <- make_mesh(
  dat,
  xy_cols = c("x", "y"),
  cutoff = 80 # min. distance between knots in X-Y units
)
# mesh$mesh$n # extract number of vertices/knots

# quick visualization of mesh
plot(mesh)
```

---

# Re-fitting the model  

Intercept only, 1 time slice  

```{r fit-model, echo=TRUE}
fit <- sdmTMB(
  n ~ 1,
  data = dat,
  mesh = mesh,
  family = truncated_nbinom2(link = "log"),
)
```

---

# Predict to grid  

* Slower, but include `se_fit = TRUE`
* Could also use `predict(..., nsim = 500)` and summarize matrix

```{r pred-fit, echo=TRUE, eval=TRUE}
# makes all combinations of x and y:
newdf <- expand.grid(
  x = seq(min(dat$x), max(dat$x), 5),
  y = seq(min(dat$y), max(dat$y), 5)
)
p <- predict(fit,
  newdata = newdf, se_fit = TRUE
)
```

---
 
# Visualizing uncertainty

* Why is uncertainty higher at vertices?  
  * Fewer neighbors, e.g. [this tutorial](https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/)
  
```{r vis-vert}
ggplot() +
  inlabru::gg(mesh$mesh) +
  geom_point(data = p, aes(x = x, y = y, col = est_se)) +
  coord_equal()
```

---
 
# Predicting outside the survey domain

* Predict into border area

```{r pred-fit2, echo=FALSE, eval=TRUE}
# makes all combinations of x and y:
newdf <- expand.grid(
  x = seq(min(dat$x) - 100, max(dat$x) + 100, 5),
  y = seq(min(dat$y) - 100, max(dat$y) + 100, 5)
)
p <- predict(fit,
  newdata = newdf, se_fit = TRUE
)
ggplot() +
  inlabru::gg(mesh$mesh) + #<<
  geom_point(data = p, aes(x = x, y = y, col = est_se)) +
  coord_equal()
```
