---
title: "Making pretty maps with sdmTMB output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making pretty maps with sdmTMB output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://pbs-assess.github.io/sdmTMB/index.html) under 'Articles'.**

```{r, include = FALSE}
sf_installed <- require("sf", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
inla_installed <- requireNamespace("INLA", quietly = TRUE)
pkgs <- sf_installed && ggplot_installed && inla_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
```

This vignette illustrates how to make gridded/raster maps of sdmTMB output (or gridded data from any source) on a base map of class sf.

```{r setup}
library(ggplot2)
library(sdmTMB)
```

First, we will download coastline data for our region. 
We will use a medium resolution version. For applied situations on finer scales, you may want to use `scale = "large"`.

```{r, eval=FALSE}
map_data <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf", country = "canada")
# Crop the polygon for plotting and efficiency:
# st_bbox(map_data) # find the rough coordinates
bc_coast <- suppressWarnings(suppressMessages(
  st_crop(map_data,
    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
```

Transform our map into UTM 9 coordinates, which is the equal-area projection we fit in:

```{r}
utm_zone9 <- 3156
bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
```

If you don't know the crs number (3156 above), you can look it up by googling it or looking on <https://epsg.io/>. For example, <https://epsg.io/3156>.

Look at our base map:

```{r}
ggplot(bc_coast_proj) + geom_sf()
```

Now we will fit a model and predict on a grid. The coordinates X and Y are already UTM zone 9 coordinates (in km).

```{r}
d <- subset(pcod, year == 2017)
pcod_spde <- make_mesh(d, c("X", "Y"), cutoff = 30)
m <- sdmTMB(density ~ 0 + depth_scaled + depth_scaled2,
  data = d, mesh = pcod_spde, family = tweedie(link = "log"))
p <- predict(m, newdata = qcs_grid)
```

We can use `sf::st_boundary()` to help find reasonable limits on our map (output not shown):

```{r, eval=FALSE}
sf::st_boundary(bc_coast_proj)
```

Finally, we will combine our gridded predictions with the base map. 
We will multiply the `X` and `Y` columns by 1000 because we worked in UTM km for model fitting (to avoid computational issues with the scale of the range paramater):

```{r, fig.width=5, fig.height=6}
ggplot(bc_coast_proj) + geom_sf() +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
  xlim(230957.7, 1157991 - 300000) +
  ylim(5366427, 6353456 - 200000) +
  scale_fill_viridis_c() +
  theme_light() +
  labs(fill = "Predicted\ndensity") +
  labs(x = "Longitude", y = "Latitude")
```
