---
title: "Fitting spatial trend models with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting spatial trend models with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://pbs-assess.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
inla_installed <- requireNamespace("INLA", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && inla_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

Using the built-in British Columbia Queen Charlotte Sound Pacific Cod dataset, we might be interested in fitting a model that describes spatially varying trends through time. The data are as follows:

- There are columns for depth and depth squared.
- Depth was centred and scaled by its standard deviation and we've included those in the data frame so that they could be used to similarly scale the prediction grid.
- The density units should be kg/km^2^.
- Here, X and Y are coordinates in UTM zone 9.

We will set up our SPDE mesh with a relatively coarse resolution so that this vignette builds quickly:

```{r}
pcod_spde <- make_mesh(pcod, c("X", "Y"), cutoff = 12)
plot(pcod_spde)
```

We will fit a model that includes a slope for 'year', an intercept spatial random field, and another random field for spatially varying slopes the represent trends over time in space (`spatial_varying` argument). Our model just estimates an intercept and accounts for all other variation through the random effects.

First, we will set up a column for time that is Normal(0, 1) to help with estimation:

```{r}
d <- pcod
d$scaled_year <- (pcod$year - mean(pcod$year)) / sd(pcod$year)
```

Now fit a model using `spatial_varying ~ 0 + scaled_year`:

(The `0 +` drops the intercept, although sdmTMB would take care of that anyways here.)

```{r}
m1 <- sdmTMB(density ~ scaled_year, data = d,
  mesh = pcod_spde, family = tweedie(link = "log"),
  spatial_varying = ~ 0 + scaled_year, time = "year",
  spatiotemporal = "off")
```

We have turned off spatiotemporal random fields for this example for simplicity, but they also could be `IID` or `AR1`.

Let's extract some parameter estimates. Look for `sigma_Z`:

```{r}
tidy(m1, conf.int = TRUE)
tidy(m1, "ran_pars", conf.int = TRUE)
```

Let's look at the predictions and estimates of the spatially varying coefficients on a grid:

```{r}
plot_map_raster <- function(dat, column = est) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(~year) +
    coord_fixed() +
    scale_fill_viridis_c()
}
```

First, we need to predict on a grid. We also need to add a column for `scaled_year` to match the fitting:

```{r}
nd <- qcs_grid
nd$scaled_year <- (nd$year - mean(pcod$year)) / sd(pcod$year)
p1 <- predict(m1, newdata = nd)
```

First let's look at the spatial trends.

We will just pick out a single year to plot since they should all be the same for the slopes. Note that these are in log space. `zeta_s` are the spatially varying coefficients.

```{r}
plot_map_raster(filter(p1, year == 2003), zeta_s_scaled_year)
```

This is the spatially varying intercept:

```{r, warning=FALSE, message=FALSE}
plot_map_raster(filter(p1, year == 2003), omega_s) + scale_fill_gradient2()
```

These are the predictions including all fixed and random effects plotted in log space.

```{r}
plot_map_raster(filter(p1, year == 2003), est)
```

And we can look at just the spatiotemporal random effects for models 2 and 3 (intercept + slope combined):

```{r}
plot_map_raster(filter(p1, year == 2003), est_rf)
```
