---
title: "Simulation test for bias in AR1 sdmTMB models"
output: html_notebook
---

Steps first executed independently and then in functions that repeat simulation and compile results
```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) 
library(sdmTMB)
library(ggplot2)
library(dplyr)
library(purrr)
library(future)
library(cowplot)
```

### Set spatial grid 
Required prior to running functions as well as in step-by-step approach

#### Create fine-scale square 50 x 50 grid to predict on
```{r}
grid <- expand.grid(X = seq(1:50), Y = seq(1:50))
```
### --OR-- 
#### Use existing spatial grid
e.g. Queen Charlotte Sound
```{r, eval=FALSE}
grid <- qcs_grid
```

## 1. Simulate 'true' data

### Output in list format along with a vector of input values
### Data parameter aguments include:
**x** A vector of x coordinates\
**y** A vector of y coordinates\
**time_steps** The number of time steps\
**ar1_fields** Logical for whether or not to include AR1 structure\
**ar1_phi** Correlation between years; should be between -1 and 1\
**sigma_O** SD of spatial process (Omega)\
**sigma_E** SD of spatiotemporal process (Epsilon)\
**kappa** Parameter that controls the decay of spatial correlation (3/kappa is roughly the distance at which points are %10 correlated)\
**phi** Observation error scale parameter\
**plot** Logical for whether or not to produce a plot\

```{r sim_args_vec}
sim_args_vec <- function(x = stats::runif(400, 0, 10), y = stats::runif(400, 0, 10),
                         time_steps = 1L, ar1_fields = FALSE, ar1_phi = 0.5,
                         sigma_O = 0.4, sigma_E = 0.3, kappa = 1.3, phi = 0.2,
                         plot = FALSE) {
  d <- sim(
    x = x, y = y, time_steps = time_steps, ar1_fields = ar1_fields, plot = plot,
    ar1_phi = ar1_phi, sigma_O = sigma_O, sigma_E = sigma_E, kappa = kappa, phi = phi
  )
  list(d, inputs = c(
    ar1_phi = ar1_phi,
    sigma_O = sigma_O,
    sigma_E = sigma_E,
    kappa = kappa,
    phi = phi
  ))
}
```

```{r}
simdat <- sim_args_vec( x = grid$X, y = grid$Y,
                        time_steps = 9,
                        plot = FALSE,
                        ar1_fields = TRUE,
                        ar1_phi = 0.5,
                        sigma_O = 0.3,
                        sigma_E = 0.3,
                        kappa = 0.15,
                        phi = 0.05
                      )
```

```{r, echo=FALSE}
g <- ggplot(simdat[[1]], aes_string("x", "y", fill = "observed")) +
              geom_raster() +
              facet_wrap(~time) +
              scale_fill_viridis_c() +
              coord_fixed()
g
```

## 2. Sub-sample from 'true' data

#### Select a random sample, evenly distributed across time periods
Total grid cells are `r nrow(simdat[[1]])` / `r length(unique(simdat[[1]]$time))` time periods = `r nrow(simdat[[1]])/length(unique(simdat[[1]]$time))`
```{r }
dat <- simdat[[1]] %>% group_by(time) %>% sample_n(500) %>% ungroup()
```

```{r, echo = FALSE}
g <- ggplot(dat, aes_string("x", "y", fill = "observed")) +
              geom_raster() +
              facet_wrap(~time) +
              scale_fill_viridis_c() +
              coord_fixed()
g
```


### --OR-- 
#### Systematically drop a chunck of data
```{r, eval = TRUE}
dat <- simdat[[1]] %>% filter((x <15|x>20) & (y<15|y>20))
```

```{r, echo = FALSE}
g <- ggplot(dat, aes_string("x", "y", fill = "observed")) +
              geom_raster() +
              facet_wrap(~time) +
              scale_fill_viridis_c() +
              coord_fixed()
g
```

## 3. Model from sub-sample

### Design mesh 
```{r}
spde <- make_spde(x = dat$x, y = dat$y, n_knots = 150)
plot_spde(spde)
```

Use sdmTMB to estimate the input parameters in our simulation.\
This model includes a spatiotemporal random field, but no fixed spatial random field.
```{r}
m <- sdmTMB(formula = observed ~ 1, 
            ar1_fields = TRUE,
            include_spatial = TRUE, 
            silent = TRUE,
            data = dat, 
            time = "time",
            family = gaussian(link = "identity"), 
            spde = spde
           )
r <- m$tmb_obj$report()
```

## 4. Compare parameter estimates with input values

Input values (from vectorized simdat list element)
```{r}
inputs <- as.data.frame(reshape::melt(simdat[[2]]))
inputs
```

Back transform parameter estimates from model report() for comparison with input values
```{r}
estimates <- c( ar1_phi = (2 * plogis(m$model$par[["ar1_phi"]]) - 1),
                sigma_O = r$sigma_O,
                sigma_E = r$sigma_E,
                kappa = exp(r$ln_kappa),
                phi = exp(r$ln_phi)
              )
estimates <- as.data.frame(reshape::melt(estimates))
estimates
```

## 5. Use model to predict for all grid points in simulated data

Replicate list of all grid points for each time period
```{r}
original_time <- sort(unique(m$data[[m$time]]))
nd <- do.call("rbind", replicate(length(original_time), grid, simplify = FALSE))
nd[[m$time]] <- rep(original_time, each = nrow(grid))
```

Run TMB with prediction turned on, and replace sample 'dat' with new grid 'nd'
```{r}
predictions <- predict(m, newdata = nd, xy_cols = c("X", "Y"))$data
glimpse(predictions)
```

## 6. Contrast true and predicted values for each point in space and time
```{r}
spatial_bias_dat <- full_join(simdat[[1]], predictions, by = c("x" = "X", "y" = "Y", "time" = "time"), suffix = c("", "_est")) %>% mutate(diff = est - real)
glimpse(spatial_bias_dat)
```

```{r}
hist(spatial_bias_dat$diff, breaks = 40)
```


### Plot spatial distribution of model estimates beside simulated true values, and the difference between them
**dataframe** Dataframe containing all simulated and predicted values to be plotted spatially\
**id** List of columns in dataframe that together identify unique observations (Default = c("x", "y", "time"))\
**values** List of values to be shown in plotted on separate panels (Default = c("real", "est", "diff"))\
**time_periods** List of time periods to be shown in plot (Default = c("1", "5", "9"))\

```{r}
plot_map_diff <- function(dataframe,
                          id = c("x", "y", "time"),
                          values = c("real", "est", "diff"),
                          time_periods = c("1", "5", "9")) {

  melted <- reshape2::melt(dataframe, id) %>% # could be replace with tidyr::gather(spatial_bias_dat, "variable", "value",... )?
    filter(variable %in% values) %>%
    filter(time %in% time_periods)

  ggplot(melted, aes_string("x", "y", fill = "value")) +
    geom_raster() +
    facet_grid(time ~ variable) +
    scale_fill_viridis_c() +
    coord_fixed()
}
plot_map_diff(spatial_bias_dat)
```

# Run simulations 'j' times for each set of parameter values

## 1. Generate dataframe with parameter arguments for multiple simulation runs
Note: This function creates a dataframe of all possible combinations,
therefore these parameters can be fixed to a single value, or vary using c(value1, value2, value3, ...)\
**ar1_phi** Correlation between years; should be between -1 and 1\
**sigma_O** SD of spatial process (Omega)\
**sigma_E** SD of spatiotemporal process (Epsilon)\
**kappa** Parameter that controls the decay of spatial correlation (3/kappa is roughly the distance at which points are %10 correlated)\
**phi** Observation error scale parameter\
**time_steps** The number of time steps\
**N** Sub-sample size = number of observations included in the sdmTMB model\
**n_knots** Number of knots for spatial process\
**j** Number of runs to be conducted for each unique combination of parameter values\

```{r}
generate_arg <- function(time_steps = 4,
                         ar1_phi = 0.5, # c(-0.85, 0.1, 0.85),
                         sigma_O = 0.3,
                         sigma_E = 0.3,
                         kappa = 0.2, # c(0.005, 0.1, 1),
                         phi = 0.1, # c(0.01, 0.1),
                         N = 300,
                         n_knots = 150,
                         j = 2L) {
  arguments <- expand.grid(
    ar1_phi = ar1_phi,
    sigma_O = sigma_O,
    sigma_E = sigma_E,
    kappa = kappa,
    phi = phi,
    N = N,
    n_knots = n_knots,
    time_steps = time_steps
  )
  nrow(arguments)
  arguments$count <- j
  arguments <- arguments[rep(seq_len(nrow(arguments)), arguments$count), ]
  arguments_apply <- dplyr::select(arguments, -count)
  nrow(arguments_apply)
  arguments_apply$iter <- 1:nrow(arguments_apply)
  arguments_apply
}
```

```{r}
args <- generate_arg(j = 10L, 
                     ar1_phi = c(0.01, 0.7))
glimpse(args)
```

## 2. Requires a function that saves a list of tibbles of:
####  [[1]] parameter inputs and estimates &
####  [[2]] real values and predicted values

**iter** Iteration id number; default is a random number; used to set.seed\
**grid** Dataframe of spatial coordinates eg. c(X, Y)\
**x** A vector of x coordinates\
**y** A vector of y coordinates\
**time_steps** The number of time steps\
**plot** Logical for whether or not to produce a plot\
**ar1_phi** Correlation between years; should be between -1 and 1\
**sigma_O** SD of spatial process (Omega)\
**sigma_E** SD of spatiotemporal process (Epsilon)\
**kappa** Parameter that controls the decay of spatial correlation (3/kappa is roughly the distance at which points are %10 correlated)\
**phi** Observation error scale parameter\
**N** Sub-sample size = number of observations included in the sdmTMB model\
**n_knots** Number of knots for spatial process#' @param formula\
**formula** Define model to be assessed with the simulated data\
**family** Set family of model to be assessed\

```{r}
sim_predictions <- function(iter = sample.int(1e3, 1), plot = FALSE,
                            grid = grid, x = grid$X, y = grid$Y,
                            time_steps = 9,
                            ar1_fields = TRUE,
                            ar1_phi = 0.5,
                            sigma_O = 0.3,
                            sigma_E = 0.3,
                            kappa = 0.05,
                            phi = 0.1,
                            N = 1000, n_knots = 150, # iter.max=1e4, eval.max=1e4,
                            formula = observed ~ 1, family = gaussian(link = "identity")) {
  set.seed(iter * 581267)

  simdat <- sim(
    plot = plot, x = x, y = y,
    time_steps = time_steps, ar1_fields = ar1_fields,
    ar1_phi = ar1_phi, sigma_O = sigma_O, sigma_E = sigma_E, kappa = kappa, phi = phi
  )

  dat <- simdat %>% group_by(time) %>% sample_n(N) %>% ungroup() # sub-sample from 'true' data
  spde <- make_spde(dat$x, dat$y, n_knots)
  plot_spde(spde)
  # browser()

  m <- sdmTMB(
    silent = TRUE,
    ar1_fields = ar1_fields,
    include_spatial = TRUE,
    data = dat, formula = formula, time = "time",
    family = family, spde = spde
  )

  r <- m$tmb_obj$report()

  estimates <- c(
    ar1_phi = (2 * plogis(m$model$par[["ar1_phi"]]) - 1),
    sigma_O = r$sigma_O,
    sigma_E = r$sigma_E,
    kappa = exp(r$ln_kappa),
    phi = exp(r$ln_phi)
  )

  inputs <- c(
    ar1_phi = ar1_phi,
    sigma_O = sigma_O,
    sigma_E = sigma_E,
    kappa = kappa,
    phi = phi
  )
  parameter <- names(inputs)
  converg <- m$model$convergence

  # replicate grid for each time period
  original_time <- sort(unique(m$data[[m$time]]))
  nd <- do.call("rbind", replicate(length(original_time), grid, simplify = FALSE))
  nd[[m$time]] <- rep(original_time, each = nrow(grid))

  # run TMB with prediction turned on but replace sample 'dat' with new grid 'nd'
  predictions <- predict(m, newdata = nd, xy_cols = c("X", "Y"))$data
  head(predictions)

  # combine true and predicted values for each point in space and time
  spatial_bias <- full_join(simdat, predictions, by = c("x" = "X", "y" = "Y", "time" = "time"), suffix = c("", "_est")) %>% mutate(diff = est - real)
  spatial_bias$converg <- converg

  unsampled <- anti_join(spatial_bias, dat)

  run <- list(par = tibble(parameter = parameter, inputs = inputs, estimates = estimates, iter = iter, converg = converg),
              allpredicted = as_tibble(spatial_bias),
              unsampled = as_tibble(unsampled))
  run
}
```

## 3. Use purrr:::pmap to loop the arguments through sim_preditions
```{r}
all_iter <- purrr::pmap(args, sim_predictions,
  grid = grid, x = grid$X, y = grid$Y,
  formula = observed ~ 1, family = gaussian(link = "identity")
)
```

# Explore and plot simulation results


If input parameter values varied, parameter and predicted dataframes need to be combined
```{r, warning= FALSE}
inputs_spread <- all_iter %>% map(~ .x[["par"]]) %>% bind_rows() %>% select( iter, parameter, inputs) %>% tidyr::spread(., key = parameter, value = c(inputs))
inputs_spread$iter <- as.factor(inputs_spread$iter)
allpredictions <- all_iter %>% map(~ .x[["allpredicted"]]) %>% bind_rows(.id = "iter") %>% left_join(., inputs_spread, by = "iter")
allunsampled <- all_iter %>% map(~ .x[["unsampled"]]) %>% bind_rows(.id = "iter") %>% left_join(., inputs_spread, by = "iter")
```


```{r}
pred.hist <- ggplot(data = allunsampled, aes(x = diff)) +
  geom_histogram(bins=40)  +
  geom_histogram(data = allpredictions, bins=40, alpha = 0.4)  +
  scale_fill_viridis_d() +
  geom_vline(xintercept = 0, linetype="dashed") +
  labs(title = "Prediction error for unsampled locations in darker colour", x = "(real - predicted)")
pred.hist
```


### Each run can be plotted spatio-temporally
```{r}
predictions_by_iter <- all_iter %>% map(~ .x[["allpredicted"]]) # %>% bind_rows(.id = "iter") %>% left_join(., inputs_spread, by = "iter") %>% split(., list("iter")) this code changes iter to character so breaks plot code below
#glimpse(predictions_by_iter)
```

```{r , eval =FALSE}
spatial.bias.plots <- purrr::map(predictions_by_iter, plot_map_diff, time_periods = c(1,2,3))
# pdf("spatial-bias-plots.pdf")
spatial.bias.plots
# dev.off()
```


### Parameter estimates from n simulations

Create tibble of parameter estimates
```{r}
params <- all_iter %>% map(~ .x[["par"]]) %>%
                       bind_rows()
```

Calculate difference between parameter value input into simulation and estimate based on sdmTMB model
```{r}
par_diff <- params %>% group_by(parameter) %>%
                       mutate(sd_est = sd(estimates), n = n()) %>%
                       group_by(iter, parameter) %>%
                       mutate(diff = (inputs - estimates), std_diff = (inputs - estimates)/sd_est) %>%
                       ungroup()
```

Histogram plotting function for error in parameter estimates\
**dataframe** Dataframe containing all simulated parameter estimates\
**x** Varible to be plotted (Default = data$std_diff)\
**xlabel** Description of variable to be plotted for use on x axis label\
**fill ** Varible used to colour bars to indicate if some estimates should be trusted more than others (Default = data$converg)\
**notes**  Description of fill choice or other caveats\
**bins ** Number of bins in histogram (Default = n/4)\

```{r}
par_error_hist <- function(dataframe = params,
                           standardize = TRUE,
                           xlabel = "Difference from input value",
                           fill = dataframe$converg,
                           notes = "if 2 colours, than some models did not converg",
                           bin_number = n/2){

  dataframe <- dataframe %>% group_by(parameter) %>%
    mutate(sd_est = sd(estimates), n = n()) %>%
    group_by(iter, parameter) %>%
    mutate(diff = (estimates - inputs), std_diff = (estimates - inputs)/sd_est) %>%
    ungroup()

  n <- dataframe$n[1]
  binn <- bin_number
  initial <- dataframe %>% group_by(parameter) %>% summarize(initial = mean(inputs), sd = round(mean(sd_est), 3))
  table <- gridExtra::tableGrob(initial, rows = NULL)

  if (standardize == TRUE) { x = dataframe$std_diff } else { x = dataframe$diff }
  
  fill <- as.factor(fill)
  plot <- ggplot(data = dataframe, aes(x = x)) +
    geom_histogram(aes(fill = fill), bins = binn)  +
    scale_fill_viridis_d() +
    geom_vline(xintercept = 0, linetype="dashed") +
    labs(title = "Simulated parameter estimates", x = xlabel) + #, caption = notes) +
    facet_wrap(~parameter, scales = "free_x") +
    theme(legend.position="none", plot.caption=element_text(size=12))

  text <- RGraphics::splitTextGrob(notes)
  ggdraw() + draw_plot(plot, x = 0, y = 0, width = 1, height = 1) +
    draw_plot(table, x = .7, y = 0.1, width = .3, height = .3) +
    draw_plot(text, x = .7, y = -.3, width = .3, height = .3)
}
par.hist <- par_error_hist(params, standardize = FALSE, xlabel = "Difference from input value")
par.hist
```

