---
title: "Spatial Modeling of Presence-Only Data with sdmTMB"
author: "Julia Indivero, Sean Anderson, Lewis Barnett, Philina English, Eric Ward, Jim Thorson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Modeling of Presence-Only Data with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
spatstat.data_installed <- require("spatstat.data", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && spatstat.data_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs,
  purl = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

While some geostatistical datasets contain information on presences and absences of species, others only include only locations where presence is observed. 
While presence-only modelling is commonly used in ecology and fisheries, there are also applications in other fields (e.g. dynamics of rare diseases). 
There are several approaches for analyzing these types of data---we illustrate two applications below, and in both utilize simulated zeros (pseudo-absences).

```{r libs, message=FALSE, warning=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(spatstat.data)
```

By adding pseudo-absences to presence-only data, we can estimate a spatial range that isn't sensitive to choice of raster or lattice resolution.

As an example dataset, we will use data on the locations of 3605 trees 
(species *Beilschmiedia pendula*) in a tropical rainforest 
from [the spatst.data package](https://rdrr.io/cran/spatstat.data/man/bei.html)

# Data

## Presence data

These data come from the `spatstat` package, and consist of point locations of 3605 trees on Barro Colorado Island (Panama). From `spatstat`, *the data files kindly supplied by Rasmus Waagepetersen*  

The data are stored in the `bei` dataframe,  

```{r dat}
dat <- data.frame(
  x = spatstat.data::bei$x,
  y = spatstat.data::bei$y
)
```

The distribution of trees is obviously not uniform and the middle of the map has very few trees.  

```{r plot-trees, fig.asp=0.5, echo=FALSE}
ggplot(dat, aes(x, y)) +
  geom_point(col = "darkblue", alpha = 0.1) +
  coord_cartesian(expand = FALSE)
```

## Generate pseudo-absences 

There are several approaches to generating pseudo-absences from these data. We first use a method called quadrature points ([Renner et al. 2015](https://doi.org/10.1111/2041-210X.12352)). For this method, we have to decide:  

1. Distribution of zeroes  
- Regularly spaced (e.g. on a grid or lattice?)
- Random  
- Higher density where environmental variability is high  (suggestion by Renner et al. 2015)

2. How many zeroes to generate?    
- Large enough so that predictive performance does not change as more are added  

Here, we will use a uniform grid strategy to create ~5000 points. 
To test whether the number of pseudo-absences is sufficient, `res` can be decreased, and model performance can be compared. 
For instance, to increase the number of pseudo-absences to ~20,000 points, we could change `res <- 5`. 
In this example, increasing the number of zeroes to ~20,000 only marginally improves model performance, so 5,000 zeroes is sufficient. 

```{r make-zeros, echo = TRUE}
res <- 10 # Determines resolution: lower value will increase number of zeroes generated

# zeros is generated on a grid for this example, but other strategies could be used
zeros <- expand.grid(
  x = seq(0, 1000, by = res),
  y = seq(0, 500, by = res)
)
```

Then, we combine the presence and pseudo-absence data, and use `$present` to separate occurrences from pseudo-absences.  

```{r bind-dat, echo=TRUE}
dat$present <- 1
zeros$present <- 0
all_dat <- rbind(dat, zeros)
```

Next we can create the mesh. 
The resolution of mesh can be changed with the `cutoff` value, which determines the minimum distance between locations in X-Y units. 
Increasing the cutoff will decrease the resolution of the mesh. 
In this example, a higher resolution mesh of `cutoff=15` marginally improved model performance over `cutoff=25`.

```{r mesh, message=FALSE, results=FALSE}
mesh <- make_mesh(
  all_dat,
  xy_cols = c("x", "y"),
  cutoff = 15 # min. distance between knots in X-Y units
)
```

The dimensions of the mesh (or number of vertices or knots) can be accessed with `mesh$mesh$n`. 
A cutoff distance of 15 yields ~ 1399 knots, and a cutoff distance of 25 yields a mesh with 624 knots. 

In the following plot, blue dots are data and red grid dots are quadrature points.

```{r mesh-viz, fig.asp=0.5, echo=FALSE}
all_dat$fpres <- as.factor(all_dat$present)
ggplot() +
  geom_point(
    data = all_dat,
    aes(x = x, y = y, col = fpres), size = 0.1, alpha = 0.3
  ) +
  coord_equal() +
  guides(col = guide_legend(title = "Present"))
```

# Model Options

## 1) Infinitely Weighted Logistic Regression (IWLR) and sdmTMB  

Several approaches exist for estimating a model with pseudo-absences. 
These are similar, but have different likelihood models. First, we can use an Infinitely Weighted Logistic Regression [Fithian & Hastie (2013)](https://doi.org/10.1214/13-AOAS667) for the model.

The first step is to calculate weights. Weights will be 1 for true occurrences and a very large number (`nW`) for pseudo-absences.  

```{r make-iwlr, echo=TRUE}
nW <- 1.0e6
all_dat$wt <- nW^(1 - all_dat$present)
```

And use sdmTMB to fit the model.

```{r iwlr-fit, echo = TRUE, warning=FALSE, message=FALSE}
fit <- sdmTMB(
  present ~ 1,
  data = all_dat,
  mesh = mesh,
  family = binomial(link = "logit"),
  weights = all_dat$wt
)
```

We can inspect the model output.

```{r inspect1, echo=TRUE}
summary(fit)
```

A criticism of this approach is that the intercept and log-likelihood can be affected by `nW` ([Renner et al. 2015](https://doi.org/10.1111/2041-210X.12352))

## 2) Downweighted Poisson Regression (DWPR)  

Another option is the Downweighted Poisson Regression, which is similar to IWLR but uses different weights and doesn't have the same arbitrary effects on intercept and likelihood  

First we re-calculate weights (note they're different than IWLR):

```{r make-dwpr, echo=TRUE}
# small values at presence locations
all_dat$wt <- 1e-6

# pseudo-absences: area per quadrature point
tot_area <- diff(range(dat$x)) * diff(range(dat$y))
n_zeros <- length(which(all_dat$present == 0))

all_dat$wt <- ifelse(all_dat$present == 1,
  1e-6, tot_area / n_zeros
)
```

Then fit the model with the new weights and a Poisson distribution

```{r fit-dwpr, echo=TRUE}
fit <- sdmTMB(
  present / wt ~ 1,
  data = all_dat,
  mesh = mesh,
  family = poisson(link = "log"),
  weights = all_dat$wt
)
```

And inspect the output

```{r inspect2, echo=TRUE}
summary(fit)
```
We can plot the random spatial effects

```{r plot-rf, echo=FALSE, fig.align="center"}
p <- predict(fit, newdata = zeros)
ggplot(p, aes(x, y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed(expand = FALSE)
```

We can predict spatial distribution both in link (log) space:

```{r plot-link, echo=FALSE, fig.align="center"}
ggplot(p, aes(x, y, fill = est)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed(expand = FALSE)
```

Or natural space:

```{r plot-norm, echo=FALSE, fig.align="center"}
ggplot(p, aes(x, y, fill = exp(est))) +
  geom_raster() +
  labs(fill = "Intensity\n(average point density)") +
  scale_fill_viridis_c(trans = "sqrt") +
  coord_fixed(expand = FALSE)
```

To evaluate the predictive performance, there are multiple options for binary data. 
AUC (Area Under the receiver operating characteristic curve) is a common metric, where values near 0.5 are essentially random, and values close to 1 indicate better predictive performance. 
We can use the package [ROCR](https://cran.rstudio.com/web/packages/ROCR/vignettes/ROCR.html).

```{r auc, echo=TRUE, eval=FALSE}
all_dat$p <- predict(fit)$est # total predictions, logit
rocr <- ROCR::prediction(exp(all_dat$p), all_dat$present)
ROCR::performance(rocr, measure = "auc")@y.values[[1]]
```

```{r echo=FALSE, eval=FALSE}
AUC <- ROCR::performance(rocr, measure = "auc")@y.values[[1]]
```
